Einstein's gravity from a polynomial affine model

### 2015-08-03 AffGR-PRD.tex
Added bibliography and small corrections

Classically, General Relativity is the most successful theory of gravitational interactions. However, it is believed to be the effective theory of a, yet to be discovered, more fundamental one. Among the problems of General Relativity we mention a few aspects related with the quantization: First, the fact that in the standard formalism of quantization General Relativity is non-renormalizable~\cite{'tHooft:1973us,'tHooft:1974bx,Deser:1974cz,Deser:1974cy}. Second, the Wheeler--DeWitt equations are not well-defined in general due to their non-polynomial dependence on the Arnowitt--Deser--Misner (or ADM) variables~\cite{Arnowitt:1959ah,Arnowitt:1960es,WheelerGeo,DeWitt:1967yk,DeWitt:1967ub,DeWitt:1967uc}. One of the most successful quantization methods of gravitational interactions ---developed from ideas by Gambini and Trias~\cite{Gambini:1980yz,Gambini:1986ew}, and Ashtekar~\cite{Ashtekar:1986yd,Ashtekar:1987gu}---, hints that General Relativity is not as complete as expected, since the Lagrangian formulation includes (at least) an extra term known as Holst term~\cite{Holst:1995pc}. On the other hand, it is well known that in the standard formulation of General Relativity the skew-symmetric part of the connection (or torsion) is assumed to be zero, albeit it is known that such assumption can be relaxed once matter is added, in particular fermionic matter.

@@

Therefore, under the premises of non-completeness and non-polynomial structure of the theory, 
%%%%%%%%% THE MODEL %%%%%%%%%
in a previous paper~\cite{Skirzewski:2014eta}, we presented a polynomial (purely) affine gravitational model in four dimensions built up entirely on the basis of full diffeomorphism invariance, and restricting ourselves to consider only naive power-counting renormalizable terms (which by no means guarantees the renormalizability of the system). We showed that its non-relativistic limit around a homogeneous and isotropic spacetime yields a Newtonian gravitational potential, despite the existence of torsion. We also highlighted the possibility of using standard methods to quantize the model, due to its polynomial structure (unlike the earliest proposals~\cite{Eddington1923math,schrodinger1950space}), and also the likelihood of avoiding the uniqueness of diffeomorphism-invariant states from Loop Quantum Gravity programme~\cite{Lewandowski:2005jk}, due to the absence of a fundamental metric.

@@

The aim of this work is to show that in the vanishing torsion sector, the field equations of the polynomial affine gravity are a generalization of the Einstein's equations.% in vacuum. In addition, we show that ---in the same sector--- under the assumption of the existence of a unique non-degenerated, covariantly constant $\binom{0}{2}$-tensor, $g$ (which could be thought as a metric), the system of a scalar field coupled with the polynomial affine gravity accepts the minimally coupled Einstein--Klein--Gordon as a solution.

The paper is organized as follows. In Sec.~\ref{model}, we present briefly the polynomial affine gravity model, and summarize the results obtained in Ref.~\cite{Skirzewski:2014eta} concerning the Newtonian limit of the model. In Sec.~\ref{rlimit}, we restrict ourselves to the sector of the theory with vanishing torsion, and find the field equations; which surprisingly are a known generalization of the Einstein's field equation. Then, in Sec.~\ref{matter} we show that under certain considerations the coupling of the gravitational model with a scalar field represents a generalization of the minimally coupled system Einstein--Klein--Gordon. Finally, in Sec.~\ref{conclusions} we summarize and discuss the possible implications of the results.

@@

%% Additionally, we studied perturbative inhomogeneous sources to the connection field equations, by considering a generic matter action.
Albeit the inclusion of matter in a non-metric spacetime is a nonstandard procedure, we assumed that the matter action would depend on the most general $\binom{2}{0}$-type symmetric tensor density, $\mathfrak{g}^{\mu \nu}$, constructed with at most quadratic terms in the available fields. This tensor density coincides with what we call the ``Eddington's metric density''~\footnote{Despite the suggestive name, this is a $\binom{2}{0}$-tensor density which does not necessarily satisfy the metric conditions. However, for the simple case of the Einstein--Hilbert action, it is in fact the inverse metric density.}, defined as the variation of the action with respect to the symmetric part of the Ricci tensor~(see Refs.~\cite{Eddington1923math,schrodinger1950space,Poplawski:2012bw}), i.e.,%% \hl{JZ: def. $R_{(\mu\nu)}$}
\begin{equation}
  \label{metric}
  \frac{\delta\ }{\delta R_{(\mu\nu)}} S[\Gamma] \equiv \sqrt{\mathsf{g}} \, \mathsf{g}^{\mu\nu} = \mathfrak{g}^{\mu\nu}.
\end{equation}
%% by assuming a matter action that depends on $\bar{g}^{\mu\nu}$.

From the geodesic equations and the linearized action on scalar perturbation, we found that only a few terms in the scalar perturbation affect the geodesic equation. The effective geodesic deviation is given by
\begin{equation}
  \label{NewtonPot}
  \Gamma^i{}_{00} = - \frac{1}{8\pi} \frac{ \partial\mathcal{L}_{\text{Matter}} }{ \partial \mathfrak{g}^{00} } \, \nabla^i \left(\frac{1}{|\vec{x}|}\right),
\end{equation}
is the usual Newtonian force induced by a massive source.

\section{\label{rlimit}Relativistic limit}
%%%%%%%%% COMPARISON GR %%%%%%%%%
Despite we obtained the correct Newtonian gravitational potential in the non-relativistic limit, see Eq.~\eqref{NewtonPot}, the post-Newtonian approximation is necessary to explain gravitational phenomena like Mercury's perihelion and the deviation of light by a gravitational source. Therefore, we should go beyond the Newtonian limit in order to compare our model with GR.

%%\hl{A comparison between our affine model and GR requires the imposition of vanishing torsion (JZ: Why?).}
Even though a comparison between our affine model and GR does not require the imposition of vanishing torsion, for the sake of simplicity, at this stage we shall focus on a sector of the theory in which the connection is compatible with a metric, and torsion-free. Notice that the vanishing torsion limit ---equivalent to take $T^{\lambda,\mu\nu} \to 0$ and $A_\mu \to 0$--- cannot be taken at the action level, and should be taken in the field equations. Interestingly, the field equations from Eq.~\eqref{4dfull} can be consistently \emph{truncated} under such requirements, because only the terms of the action linear in these fields will be relevant, i.e.,
\begin{equation*}
  %\label{eff-action}
  S_{\text{eff}} = \int\dn{4}{x} \Big( C_1\, R_{\mu\rho}{}^{\mu}{}_\nu \nabla_\sigma T^{\nu,\rho\sigma} + C_2 \, R_{\mu\nu}{}^{\rho}{}_\rho \nabla_\sigma T^{\sigma,\mu\nu} \Big),
\end{equation*}
and the only nontrivial field equation is the one for the Curtright, $T^{\nu,\mu\rho}$,
\begin{equation*}
  \nab{[\rho} R_{\mu]\lambda}{}^{\lambda}{}_{\nu} + \kappa \nab{\nu} R_{\mu\rho}{}^\lambda{}_\lambda = 0,
\end{equation*}
with $\kappa$ a constant related with the couplings of the model. In the considered sector the connection is compatible with a metric, therefore, the second term vanishes and the gravitational equations are
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} = 0.
  \label{SimpleEOM}
\end{equation}

Equation~\eqref{SimpleEOM} is a generalization of a condition known in Riemannian geometry as covariantly constant Ricci curvature ---aka parallel Ricci curvature---, \mbox{$\nab{\rho} R_{\mu\nu} = 0$.} For all Einstein manifolds, whose Ricci tensor is proportional to the metric, \mbox{$R_{\mu\nu} \propto g_{\mu\nu}$,} satisfy the parallel Ricci condition, and therefore every vacuum solution to the Einstein's equations solves the (torsion-free) field equations of our model. Consequently, the fact that the non-relativistic limit of the gravitational potential in Eq.~\eqref{NewtonPot} yields a Newtonian potential seems clearer, and we can argue that even the post-Newtonian corrections coming from GR are present in the chosen scenario of our model. As an additional comment, the parallel Ricci curvature assures that although the manifold is not in general a Einstein manifold, the metric has to be locally a product of Einstein metrics~\cite{Besse}.

@@

For a Riemannian manifold $(\Mi,g)$ the Eq.~\eqref{SimpleEOM} is related through the first Bianchi identity to the harmonic curvature condition~\cite{bourguignon1981varietes},
\begin{equation}
  \label{harm-curv}
  \nabla_\lambda R_{\mu\nu}{}^\lambda{}_\rho = 0,
\end{equation}
and it can be shown that a manifold with harmonic curvature is equivalent (in four dimensions) to a manifold with harmonic Weyl tensor~\cite{Berger:1969} and constant scalar curvature, or in other words the Ricci tensor is a Codazzi tensor~\footnote{A Codazzi tensor is a symmetric $(0,2)$-type tensor, $T$, satisfying the condition \mbox{$D_X T(Y,Z) = D_Y T(X,Z)$~\cite{Derdzinski01071983}.}}. For prove of these equivalences, see Refs.~\cite{Derdzinski:1985,Besse}.

@@

Summarizing, we have shown that the  \emph{purely affine gravity} constructed in Ref.~\cite{Skirzewski:2014eta} possesses a well-defined torsionless limit, which generalizes the equations of motion  from General Relativity, and its field equations are equivalent to those obtained from a Yang--Mills theory for the spacetime connection, or Stephenson--Kilmister--Yang theory. Consequently, all gravitational effects described by standard General Relativity are contained in our model, even though the torsion field is kept to zero. Regarding this aspect, we would like to emphazise that there is enough liberty to start considering new cosmological effects coming from both nonvanishing torsion and non-metric connections with vanishing torsion, beside the results reported in Ref.~\cite{Chen:2013kia,Chen:2013ota}.
Moreover, our gravitational model coupled to a scalar field through the \emph{tensor density} $\mathfrak{g}^{\mu\nu}$ ---in the vanishing torsion consideration---, generalizes GR minimally coupled to a massless scalar field. In a forthcoming paper, we will consider more general matter content (unrelated to the work in Ref.~\cite{Cook:2008mx}), and some properties of the field equations which differ from those of General Relativity.

@@

%%%%%%%%% BIBLIOGRAPHY %%%%%%%%%
\bibliographystyle{apsrev4-1}
\bibliography{References.bib}

>>>>

Classically, General Relativity is the most successful theory of gravitational interactions. However, it is believed to be the effective theory of a, yet to be discovered, more fundamental one. Among the problems of General Relativity we mention a few aspects related with the quantization: First, the fact that in the standard formalism of quantization General Relativity is non-renormalizable~\cite{'tHooft:1973us,'tHooft:1974bx,Deser:1974cz,Deser:1974cy}. Second, the Wheeler--DeWitt equations are not well-defined in general due to their nonpolynomial dependence on the Arnowitt--Deser--Misner (or ADM) variables~\cite{Arnowitt:1959ah,Arnowitt:1960es,WheelerGeo,DeWitt:1967yk,DeWitt:1967ub,DeWitt:1967uc}. One of the most successful quantization methods of gravitational interactions ---developed from ideas by Gambini and Trias~\cite{Gambini:1980yz,Gambini:1986ew}, and Ashtekar~\cite{Ashtekar:1986yd,Ashtekar:1987gu}---, hints that General Relativity is not as complete as expected, since the Lagrangian formulation includes (at least) an extra term known as Holst term~\cite{Holst:1995pc}. On the other hand, it is well known that in the standard formulation of General Relativity the skew-symmetric part of the connection (or torsion) is assumed to be zero, albeit it is known that such assumption can be relaxed once matter is added, in particular fermionic matter~\cite{Kibble:1961ba}.

@@

Therefore, under the premises of incompleteness and non-polynomial structure of the theory, 
%%%%%%%%% THE MODEL %%%%%%%%%
in a previous paper~\cite{Skirzewski:2014eta}, we presented a polynomial (purely) affine gravitational model in four dimensions built up entirely on the basis of full diffeomorphism invariance, and restricting ourselves to consider only naive power-counting renormalizable terms (which by no means guarantees the renormalizability of the system). We showed that its non-relativistic limit around a homogeneous and isotropic spacetime yields a Newtonian gravitational potential, despite the existence of torsion. We also highlighted the possibility of using standard methods to quantize the model, due to its polynomial structure (unlike the earliest proposals~\cite{Eddington1923math,schrodinger1950space}), and also the likelihood of avoiding the uniqueness of diffeomorphism-invariant states from Loop Quantum Gravity programme~\cite{Lewandowski:2005jk}, due to the absence of a fundamental metric, or in other words, the absence of flux operators.

@@

The aim of this work is to show that in the vanishing torsion sector, the field equations of the polynomial affine gravity are a generalization of the Einstein's equations. % in vacuum. In addition, we show that ---in the same sector--- under the assumption of the existence of a unique non-degenerated, covariantly constant $\binom{0}{2}$-tensor, $g$ (which could be thought as a metric), the system of a scalar field coupled with the polynomial affine gravity accepts the minimally coupled Einstein--Klein--Gordon as a solution.
The paper is organized as follows. In Sec.~\ref{model}, we present briefly the polynomial affine gravity model, and summarize the results obtained in Ref.~\cite{Skirzewski:2014eta} concerning the Newtonian limit of the model. In Sec.~\ref{rlimit}, we restrict ourselves to the sector of the theory with vanishing torsion, and find the field equations; which surprisingly are a known generalization of the Einstein's field equation. Then, in Sec.~\ref{matter} we show that under certain considerations the coupling of the gravitational model with a scalar field represents a generalization of the minimally coupled system Einstein--Klein--Gordon. Finally, in Sec.~\ref{conclusions} we summarize and discuss the possible implications of the results.

@@

%% Additionally, we studied perturbative inhomogeneous sources to the connection field equations, by considering a generic matter action.
Albeit the inclusion of matter in a nonmetric spacetime is a nonstandard procedure, we assumed that the matter action would depend on the most general $\binom{2}{0}$-type symmetric tensor density, $\mathfrak{g}^{\mu \nu}$, constructed with at most quadratic terms in the available fields. A simple \emph{dimensional analysis} shows that this tensor density contains the same terms as what we call the ``Eddington's metric density''~\footnote{Despite the suggestive name, this is a $\binom{2}{0}$-tensor density which does not necessarily satisfy the metric conditions. However, for the simple case of the Einstein--Hilbert action, it is in fact the inverse metric density.}, defined as the variation of the action with respect to the symmetric part of the Ricci tensor~(see Refs.~\cite{Eddington1923math,schrodinger1950space,Poplawski:2012bw}), i.e.,%% \hl{JZ: def. $R_{(\mu\nu)}$}
\begin{equation}
  \label{metric}
  \frac{\delta\ }{\delta R_{(\mu\nu)}} S[\Gamma] \equiv \sqrt{\mathsf{g}} \, \mathsf{g}^{\mu\nu} = \bar{g}^{\mu\nu}.
\end{equation}
%% by assuming a matter action that depends on $\bar{g}^{\mu\nu}$.

From the geodesic equations and the linearized action on scalar perturbation, we found that only a few terms in the scalar perturbation affect the geodesic equation~\footnote{We cross-check our result using the software Cadabra~\cite{Peeters2007550,peeters2007symbolic,Peeters:2007wn}.}. The effective geodesic deviation is given by
\begin{equation}
  \label{NewtonPot}
  \Gamma^i{}_{00} = - \frac{1}{8\pi} \frac{ \partial\mathcal{L}_{\text{Matter}} }{ \partial \mathfrak{g}^{00} } \, \nabla^i \left(\frac{1}{|\vec{x}|}\right),
\end{equation}
is the usual Newtonian force induced by a massive source.

\section{\label{rlimit}Relativistic limit}
%%%%%%%%% COMPARISON GR %%%%%%%%%
Despite we obtained the correct Newtonian gravitational potential in the non-relativistic limit, see Eq.~\eqref{NewtonPot}, the post-Newtonian approximation is necessary to explain gravitational phenomena like Mercury's perihelion and the deviation of light by a gravitational source. Therefore, we should go beyond the Newtonian limit in order to compare our model with General Relativity.

%%\hl{A comparison between our affine model and GR requires the imposition of vanishing torsion (JZ: Why?).}
Even though a comparison between the models does not require the imposition of vanishing torsion, for the sake of simplicity, at this stage we shall focus on a sector of the theory in which the connection is torsion-free. Notice that the vanishing torsion limit ---equivalent to take $T^{\lambda,\mu\nu} \to 0$ and $A_\mu \to 0$--- cannot be taken at the action level, and should be taken in the field equations. Interestingly, the field equations from Eq.~\eqref{4dfull} can be consistently \emph{truncated} under such requirements. Since only the terms of the action linear in these fields will be relevant, one can consider the effective action linear in the torsion's fields, i.e.,
\begin{equation}
  %\label{eff-action}
  S_{\text{eff}} = \int\dn{4}{x} \Big( C_1\, R_{\lambda\mu}{}^{\lambda}{}_\nu \nabla_\rho %T^{\nu,\rho\sigma}
  + C_2 \, R_{\mu\rho}{}^{\lambda}{}_\lambda \nabla_\nu \Big) T^{\nu,\mu\rho} ,
\end{equation}
and the only nontrivial field equation after the limit will be the one for the Curtright, $T^{\nu,\mu\rho}$,
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} + \kappa \nab{\nu} R_{\mu\rho}{}^\lambda{}_\lambda = 0,
\end{equation}
with $\kappa$ a constant related with the original couplings of the model. We assume that the connection is compatible with volume form, i.e., it is equi-affine~\cite{nomizu1994affine,MO-Bryant02}.  The equi-affine condition assures that the Ricci tensor of the connection is symmetric, and the contraction of the last indices vanishes, thus the second term in the field equation is absent and the gravitational equations are
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} = 0.
  \label{SimpleEOM}
\end{equation}

Equation~\eqref{SimpleEOM} is a generalization of a condition known in Riemannian geometry as covariantly constant Ricci curvature ---aka parallel Ricci curvature---, \mbox{$\nab{\rho} R_{\mu\nu} = 0$.}  All Einstein manifolds, whose Ricci tensor is proportional to the metric, \mbox{$R_{\mu\nu} \propto g_{\mu\nu}$,} satisfy the parallel Ricci condition due to the metricity condition, and therefore every vacuum solution to the Einstein's equations solves the (torsion-free) field equations of our model. Consequently, the fact that the non-relativistic limit of the gravitational potential in Eq.~\eqref{NewtonPot} yields a Newtonian potential seems clearer, and we can argue that even the post-Newtonian corrections coming from General Relativity are present in the chosen scenario of our model. As an additional comment, the parallel Ricci curvature assures that although the manifold is not in general a Einstein manifold, the metric has to be locally a product of Einstein metrics~\cite{Besse}.

@@

The Eq.~\eqref{SimpleEOM} is related through the second Bianchi identity to the harmonic curvature condition~\cite{bourguignon1981varietes},
\begin{equation}
  \label{harm-curv}
  \nabla_\lambda R_{\mu\nu}{}^\lambda{}_\rho = 0,
\end{equation}
and it can be shown that a manifold with harmonic curvature is equivalent (in four dimensions) to a manifold with harmonic Weyl tensor~\cite{Berger:1969} and constant scalar curvature, or in other words the Ricci tensor is a Codazzi tensor~\footnote{A Codazzi tensor is a symmetric $(0,2)$-type tensor, $T$, satisfying the condition \mbox{$D_X T(Y,Z) = D_Y T(X,Z)$~\cite{Derdzinski01071983}.}}. For prove of these equivalences, see Refs.~\cite{Derdzinski:1985,Besse}.

@@

Summarizing, we have shown that the  \emph{purely affine gravity} constructed in Ref.~\cite{Skirzewski:2014eta} possesses a well-defined torsionless limit, which generalizes the equations of motion  from General Relativity, and its field equations are equivalent to those obtained from a Yang--Mills theory for the spacetime connection, or Stephenson--Kilmister--Yang theory. Consequently, all gravitational effects described by standard General Relativity are contained in our model, even though the torsion field is kept to zero. Regarding this aspect, we would like to emphazise that there is enough liberty to start considering new cosmological effects coming from both nonvanishing torsion and nonmetric connections with vanishing torsion, beside the results reported in Ref.~\cite{Chen:2013kia,Chen:2013ota}.
Moreover, our gravitational model coupled to a scalar field through the \emph{tensor density} $\mathfrak{g}^{\mu\nu}$ ---in the vanishing torsion consideration---, generalizes GR minimally coupled to a massless scalar field. In a forthcoming paper, we will consider more general matter content (unrelated to the work in Ref.~\cite{Cook:2008mx}), and some properties of the field equations which differ from those of General Relativity.

@@

%%%%%%%%% BIBLIOGRAPHY %%%%%%%%%
%%\bibliographystyle{apsrev4-1}
\bibliography{References.bib}
